---
title: "Lab 6"
author: "J. Kuhlman"
date: "2022-10-31"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(tidyverse)
library(here)
library(janitor)
library(ggpubr)
library(car)
```

# Lab 6

## Part 1: R Markdown Techniques
Moved to lecture on Wed so we can focus on understanding stat output today

## Part 2: Plots and ANOVA Assumptions
```{r, message=FALSE}
penguins<-read_csv((here("data","penguins.csv")))
```

Subset of the penguin data
```{r}
penguins_males<-penguins %>% 
  filter(sex=="male") %>% 
  select(species, body_mass_g)
```

*Visual Inspection*
```{r}
#break out the data
adelie <- penguins_males %>% filter(species=="Adelie")
chinstrap <- penguins_males %>% filter(species=="Chinstrap")
gentoo <- penguins_males %>% filter(species=="Gentoo")

#qqplots
qq_adelie <- ggplot(adelie, aes(sample=body_mass_g))+geom_qq()+
  stat_qq()+stat_qq_line()+
  theme_bw() + labs(title="Adelie")
qq_adelie

qq_chinstrap <- ggplot(chinstrap, aes(sample=body_mass_g))+geom_qq()+
  stat_qq()+stat_qq_line()+
  theme_bw() + labs(title="Chinstrap")
qq_chinstrap

qq_gentoo <- ggplot(gentoo, aes(sample=body_mass_g))+geom_qq()+
  stat_qq()+stat_qq_line()+
  theme_bw() + labs(title="Gentoo")
qq_gentoo
```
*Combine three figures into one*
```{r}
#as a row of figures
ggarrange(qq_adelie, qq_chinstrap, qq_gentoo, labels=c("A","B","C"), ncol=3, nrow=1)

#as a column of figures
ggarrange(qq_adelie, qq_chinstrap, qq_gentoo, labels=c("A","B","C"), ncol=1, nrow=3)

#there are ways to customize the look of these - struggle with data and formatting for clarity!
```

### Critical Test for Normality
```{r}
shapiro.test(adelie$body_mass_g)
#underlying hypothesis: data are normally distributed; alternative hypothesis: not normal. With alpha of .05, and p-value of .416, we fail to reject the null, and data are normally distributed

shapiro.test(chinstrap$body_mass_g)

shapiro.test(gentoo$body_mass_g)
```

Critical part of ANOVA is also equal variance:
*test of equal variances*
```{r}
#levene's test:
leveneTest(body_mass_g~as.factor(species), data=penguins_males) #we fail to reject that the variances of all 3 penguins are about the same
```

## Part 3a: ANOVA
```{r}
m_penguin_species_results <- aov(body_mass_g~as.factor(species), data=penguins_males)
m_penguin_species_results #this just runs the calcs, but doesn't include the p-val. for that, call the results:
summary(m_penguin_species_results)
#conclude means across the three species are not all the same b/c less than 0.05 threshold
```

## Part 3b: Multiple Comparisons
```{r}
TukeyHSD(m_penguin_species_results)
#looks at the different p-values, pair-wise comparison between all possible combinations. Tukey says you have 3 comparisons, we'll adjust the p-value (p/3) to account for multiple combos
#.299 - means are relatively the same; gentoo and adelie are different, gentoo and chinstrap are different btwn p vals are so small. Let the stats guide, then verify. 
```

*New Plot to Show ANOVA*
```{r}
ggline(penguins_males, x="species", y="body_mass_g", main="Means and 95% CI for three species of penguins", add = c("mean_ci"), color="dark gray", ylab = "Body Mass (g)", xlab = "Species", point.color = "black")
```

## Part 4: Non parametric tests

```{r}
# Kruskal test
kruskal.test(body_mass_g~as.factor(species), data=penguins_males)
#h naught: medians are the same. p value is very small, so we conclude that we reject the null hypothesis that the medians are all the same
```
*pairwise comparisons*
```{r}
pairwise.wilcox.test(penguins_males$body_mass_g, penguins_males$species, p.adjust.method = "bonferroni")
#bonferroni is simple adjustment. divide by 3. 
#we conclude we fail to reject adelie and chinstrap p-values. Use non-parametrics when you have skewed data, median is more effective than mean in skewed comparisons
```

*box plot*
```{r}
ggboxplot(penguins_males, x = "species", y = "body_mass_g")
```

## Part 5: Intro to Linear Regression
*data*
```{r}
trees_mod<-read.csv(here("data", "trees_mod.csv"))

trees<-trees_mod %>% clean_names()

head(trees)
```

### Scatterplot of data
```{r}
tree_scatter<-ggplot(trees, aes(x=height, y=girth))+geom_point()+
  labs(x="Height (ft)", y="DBH (in)") + xlim(60,90) + ylim(0,25) + theme_bw() #DBH "Diamerter at Breast Height"
tree_scatter
```

*linear models*
```{r}
model_1 <- lm(girth~height, data=trees)
model_1

summary(model_1)
#height p val shows that there's a significant slope to the data. This
```

*Plot the Linear Model*

```{r}
tree_scatter<-ggplot(trees, aes(x=height, y=girth))+geom_point()+
  labs(x="Height (ft)", y="DBH (in)") + xlim(60,90) + ylim(0,25) + theme_bw() + #DBH "Diameter at Breast Height"
  geom_smooth(method="lm")
tree_scatter
```

